The Unicode Cookbook \linebreak for Linguists

### 2016-08-18 book/unicode.tex
update general files

% Everything following a % is ignored
% Some lines start with %. Remove the % to include them

@@

% put all additional commands you need in the 
% following files. If you do not know what this might 
% mean, you can safely ignore this section
\usepackage{verbatim}
\input{localmetadata.tex}
\input{localpackages.tex}
\input{localhyphenation.tex}
\input{localcommands.tex}
\bibliography{localbibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%                                              %%%
%%%             Frontmatter                      %%%
%%%                                              %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\maketitle
\frontmatter
% %% uncomment if you have preface and/or acknowledgements
% \include{chapters/preface}
% \include{chapters/acknowledgments}
% \include{chapters/abbreviations}
\tableofcontents
\mainmatter%

@@

\section{Available implementations}

To illustrate the practical application of orthography profiles as defined in
Chapter \ref{orthography-profiles}, we have implemented two different versions
of the specifications presented there: one in Python and one in R. These two
implementations have rather different implementation histories, and we are not
100\% sure that they will in all situations give the same results. Also the
perfomance with larger datasets differs, and the code is not always as clean as
we would like it to be. In sum, the two implementations should be considered as
`proof of concept' and not as the final word on the practical application of the
specifications above. In our own experience, the current implementations are
sufficiently fast and stable to be useful for academic practice (e.g. checking
data consistency, or analyzing and transliterating small to medium sized data
sets), but they should probably better not be used for full-scale industry
applications.

This chapter will introduce the implementations, and give practical step-by-step 
guidelines for installing them and using them. Various simple, and sometimes 
somewhat abstract, examples will be discussed to show the different options 
available, and to illustrate the intended usage of orthography profiles. In the 
next chapter we will then discuss a few full-scale examples to illustrate the 
application and the usefulness of orthography profiles.

@@

\noindent The version available through CRAN is the most recent stable version.
To obtain the most recent bug-fixes and experimental additions, please use the
development versions, which is available on
GitHub.\footnote{\url{http://github.com/cysouw/qlcData}} This development
version can be easily installed using the github-install helper software from the
\texttt{devtools} package.

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# download and install helper software}
\hlkwd{install.packages}\hlstd{(}\hlstr{"devtools"}\hlstd{)}
\hlcom{# install the qlcData package from GitHub}
\hlstd{devtools}\hlopt{::}\hlkwd{install_github}\hlstd{(}\hlstr{"cysouw/qlcData"}\hlstd{)}
\hlcom{# load the software, so it can be used}
\hlkwd{library}\hlstd{(qlcData)}
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\section{Error reporting}
\label{error-reporting}

In the following example, it looks as if we have two identical strings,
\texttt{AABB}. However, this is just an surface-impression delivered by the
current font, which renders Latin and Cyrillic capitals identically. We can
identify this problem when we produce an orthography profile from the strings.
Using here the R implementation of orthography profiles, we first assign the two
strings to a variable \texttt{test}, and then produce an orthography profile
with the function \texttt{write.profile}. As it turns out, some of the letters 
are Cyrillic.

@@

\noindent The working of error-message reporting can also nicely be illustrated
with this example. Suppose we made an orthography profile with just the two
Latin letters <A> and <B> as possible graphemes, then this profile would not be
sufficient to tokenize the strings. There are graphemes in the data that are not
in the profile, so the tokenization produces an error, which can be used to fix
the encoding. In the example below, we can see that the Cyrillic encoding is
found in the second string of the \texttt{test} input.

@@

The function \texttt{write.profile} can be used to prepare a skeleton for an
orthography profile from some data. The preparation of an orthography profile
from some data might sound like a trivial problem, but actually there are
various different ways in which strings can be separated into graphemes by
\texttt{write.profile}. Consider the following example string. The default
settings of \texttt{write.profile} separates the string into Unicode graphemes
according to grapheme clusters (`user-perceived characters', see
Section~\ref{the-unicode-approach}).

@@

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Thu Jun 30 18:15:48 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 1 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ú & 1 & U+0055, U+0301 & LATIN CAPITAL LETTER U, COMBINING ACUTE ACCENT \\ 
  Ù & 1 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Ù & 1 & U+0055, U+0300 & LATIN CAPITAL LETTER U, COMBINING GRAVE ACCENT \\ 
  Û & 1 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
  Û & 1 & U+0055, U+0302 & LATIN CAPITAL LETTER U, COMBINING CIRCUMFLEX ACCENT \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 1 (default settings, splitting grapheme clusters)} 
\label{tab:profile1}
\end{table}

\noindent Some of these graphemes are single codepoints, others are combinations
of two codepoints. By specifying the splitting separator as empty, it is
possible to split the string into Unicode codepoints.

@@

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Thu Jun 30 18:15:48 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
́ & 1 & U+0301 & COMBINING ACUTE ACCENT \\ 
  ̀ & 1 & U+0300 & COMBINING GRAVE ACCENT \\ 
  ̂ & 1 & U+0302 & COMBINING CIRCUMFLEX ACCENT \\ 
  U & 3 & U+0055 & LATIN CAPITAL LETTER U \\ 
  Ú & 1 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 1 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 1 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 2 (splitting by codepoints)} 
\label{tab:profile2}
\end{table}

\noindent Some characters look identical, although they are encoded differently. 
Unicode offers different ways of normalization, which can be invoked here as 
well (see Section~\ref{pitfall-canonical-equivalence}).

@@

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Thu Jun 30 18:15:48 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 2 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 2 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 3 (splitting by NFC codepoints)} 
\label{tab:profile3}
\end{table}

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Thu Jun 30 18:15:48 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
́ & 2 & U+0301 & COMBINING ACUTE ACCENT \\ 
  ̀ & 2 & U+0300 & COMBINING GRAVE ACCENT \\ 
  ̂ & 2 & U+0302 & COMBINING CIRCUMFLEX ACCENT \\ 
  U & 6 & U+0055 & LATIN CAPITAL LETTER U \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 4 (splitting by NFD codepoints)} 
\label{tab:profile4}
\end{table}

\noindent It is important to realize that for Unicode grapheme definitions, NFC 
and NFD normalization are equivalent. This can be shown by normalizing the 
example in either NFC or NFD, but using default separation in 
\texttt{write.profile}. To be precise, default separation means setting \texttt{sep = NULL} 
but that has not be added explicitly.

@@

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Thu Jun 30 18:15:48 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+0055, U+0301 & LATIN CAPITAL LETTER U, COMBINING ACUTE ACCENT \\ 
  Ù & 2 & U+0055, U+0300 & LATIN CAPITAL LETTER U, COMBINING GRAVE ACCENT \\ 
  Û & 2 & U+0055, U+0302 & LATIN CAPITAL LETTER U, COMBINING CIRCUMFLEX ACCENT \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 5 (splitting by graphemes after NFD)} 
\label{tab:profile5}
\end{table}

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Thu Jun 30 18:15:49 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 2 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 2 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 6 (splitting by graphemes after NFC)} 
\label{tab:profile6}
\end{table}

\noindent Note that these different profiles can also be produced using the bash 
executable \texttt{writeprofile} (see \ref{implementations} for notes on how to 
install the bash executable). Exactly this example is also included in the help 
file of the executable.

@@

\begin{verbatim}
## USAGE: 
##   writeprofile [-hie] [-s SEP] [-n NORMALIZE] [<STRINGS>]
## 
## DESCRIPTION:
##   Using the function write.profile() from the R-package qlcData to prepare 
##   a skeleton for an orthography profile to be used with tokenize. Details:
##   http://www.rdocumentation.org/packages/qlcData/functions/readwrite_orthography
## 
##   STRINGS can be piped through. The results is a tab-delimited file for manual
##   editing. Note that the R-code allows for even more options, not all options 
##   are made available here for ease of use.
## 
## OPTIONS:
##   -h, --help      Showing this help text
##   -i, --info      Add columns with Unicode information to profile
##   -e, --editing   Add empty columns for easy editing of profile
##   -s SEP          Separator used to separate the strings
##                     Often useful is option UNI to split by unicode codepoints
##                     defaults to unicode character definition [default: NULL]
##   -n NORMALIZE    Normalization used to make profile. Possible options: NFD, NFC
##                     defaults to no normalization [default: NULL]
##                     
## EXAMPLES:
##   Note the differences between the following variants:
## 
##     example='ÙÚÛÙÚÛ'
##     echo -e $example
##     echo -e $example | writeprofile -i
##     echo -e $example | writeprofile -i -s UNI
##     echo -e $example | writeprofile -i -s UNI -n NFC
##     echo -e $example | writeprofile -i -s UNI -n NFD
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Thu Jun 30 18:15:49 2016
\begin{table}[htb]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & UnicodeName \\ 
  \midrule
sh &  &  &  \\ 
  ch &  &  &  \\ 
  sch &  &  &  \\ 
  ie &  &  &  \\ 
  oo &  &  &  \\ 
  a & 2 & U+0061 & LATIN SMALL LETTER A \\ 
  c & 2 & U+0063 & LATIN SMALL LETTER C \\ 
  e & 3 & U+0065 & LATIN SMALL LETTER E \\ 
  f & 1 & U+0066 & LATIN SMALL LETTER F \\ 
  h & 4 & U+0068 & LATIN SMALL LETTER H \\ 
  i & 3 & U+0069 & LATIN SMALL LETTER I \\ 
  m & 4 & U+006D & LATIN SMALL LETTER M \\ 
  o & 2 & U+006F & LATIN SMALL LETTER O \\ 
  p & 2 & U+0070 & LATIN SMALL LETTER P \\ 
  s & 4 & U+0073 & LATIN SMALL LETTER S \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Manually edited profile skeleton} 
\label{tab:profile_skeleton1}
\end{table}

For quick solutions, it is also possible to leave out the Unicode information in
the profile skeleton by using the option \texttt{info = FALSE}. It is also
possible not to use a separate file at all, but process everything within R. In
simply situations this is often useful (see below), but in general we prefer to
handle everything through a separately saved orthography profile. This profile
often contains highly useful information that is nicely coded and saved inside
this one file, and can thus be easily distributed and shared. Doing the same as
above completely within R might look as follows:

@@

This will lead to the following four files being written:

   \item \textsc{result_strings.tsv}:\\ A tab-separated file with the original
         and the tokenized/transliterated strings

   \item \textsc{result_profile.tsv}:\\ A tab-separated file with the
         graphemes with added frequencies of occurence in the data. The lines
         in the file is re-ordered according to the order that resulted from the
         ordering specifications.

   \item \textsc{result_errors.tsv}:\\ A tab-separated file with all original
         strings that contain unmatched parts. Unmatched parts are indicated
         with the character as specified with the option \texttt{missing}. By
         default the character \textsc{double question mark} < ⁇ > at
         \uni{2047} is used. When there are no errors, then this file is 
         absent.

    \item \textsc{result_missing.tsv}:\\ A tab-separated file with the graphemes
          that are missing from the original orthography profile, as indicated in
          the errors. When there are no errors, then this file is absent.

@@

Not yet everything is correct in this graphemic parsing. The sequence <sh> in
`mishap' should not be a digraph, and likewise the sequence <sch> in `mischief'
should of course be separated into <s> and <ch>. One of the important issues to
get the graphemic parsing right is the order in which graphemes are parsed. For
example, currently the grapheme <sch> is parsed before the grapheme <ch>,
leading to <m\ i\ sch\ ie\ f> instead of the intended <m\ i\ s\ ch\ ie\ f>. The
reason that <sch> is parsed before <ch> is that by default longer graphemes are
parsed before shorter ones. Our experience is that in most cases this is
expected behaviour. You can change the ordering by specifying the option
\texttt{ordering}. Setting this option to \texttt{NULL} results in no
preferential ordering, i.e. the graphemes are parsed in the order of the
profile, from top to bottom. Now `mischief' is parsed correctly, but `scheme' is
wrong. So ordering is not the solution in this case.

@@

   \item \textsc{size}:\\
         order the lines in the profile by the size of the
         grapheme, largest first. Size is measured by number of Unicode
         characters after normalization as specified in the option
         \texttt{normalize}. For example, <é> has a size of 1 with
         \texttt{normalize = "NFC"}, but a size of 2 with
         \texttt{normalize = "NFD"}.

   \item \textsc{context}:\\
         order the lines by whether they have any context
         specified (see next section). Lines with context coming first. Note
         that this only works when the option \texttt{regex = TRUE} is also chosen.

  \item \textsc{reverse}:\\
        order the lines from bottom to top.

  \item \textsc{frequency}:\\
         order the lines by the frequency with which they
         match in the specified strings before tokenization, least frequent
         coming first. This frequency of course depends crucially on the
         available strings, so it will lead to different orderings when applied
         to different data. Also note that this frequency is (necessarily)
         measured before graphemes are identified, so these ordering frequencies
         are not the same as the final frequencies shown in the output.
         Frequency of course also strongly differs on whether context is used
         for the matching through \texttt{regex = TRUE}.

@@

By specifying more than one ordering, these orderings are used to break ties,
e.g. the default setting \texttt{ordering = c("size", "context", "reverse")}
will first order by size, and for those with the same size, it will order by
whether there is any context specified or not. For lines that are still tied
(i.e. the have the same size and both/neither have context) the order will be
reversed compared to the order as attested in the profile. Reversing order can
be useful because hand-written profiles tend to put general rules before
specific rules, which mostly should be applied in reverse order.

@@

To document the process of graphemic tokenization, it is strongly encouraged to

>>>>

@@

% put all additional commands you need in the 
% following files. If you do not know what this might 
% mean, you can safely ignore this section

\usepackage{verbatim}
\input{localmetadata.tex}
\input{localpackages.tex}
\input{localhyphenation.tex}
\input{localcommands.tex}
\bibliography{localbibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%                                              %%%
%%%             Frontmatter                      %%%
%%%                                              %%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}
\maketitle
\frontmatter
% \include{chapters/preface}
% \include{chapters/acknowledgments}
% \include{chapters/abbreviations}
\tableofcontents
\mainmatter%

@@

\section{Available implementations}
\label{available-implementations}

To illustrate the practical application of orthography profiles as defined in
the previous Chapter \ref{orthography-profiles}, we have implemented two
different versions of the specifications presented there: one in Python and one
in R. These two implementations have rather different implementation histories,
and we are not 100\% sure that they will in all situations give the same
results. Also the perfomance with larger datasets differs, and the code is not
always as clean as we would like it to be. In sum, the two implementations
should be considered as `proof of concept' and not as the final word on the
practical application of the specifications above. In our own experience, the
current implementations are sufficiently fast and stable to be useful for
academic practice (e.g. checking data consistency, or analyzing and
transliterating small to medium sized data sets), but they should probably
better not be used for full-scale industry applications.

This chapter will introduce the implementations, and give practical step-by-step 
guidelines for installing them and using them. Various simple and sometimes 
somewhat abstract examples will be discussed to show the different options 
available, and to illustrate the intended usage of orthography profiles. In the 
next chapter we will then discuss a few full-scale examples to illustrate the 
application and the usefulness of orthography profiles.

@@

\noindent The version available through CRAN is the latest stable version.
To obtain the most recent bug-fixes and experimental additions, please use the
development version, which is available on
GitHub.\footnote{\url{http://github.com/cysouw/qlcData}} This development
version can be easily installed using the github-install helper software from the
\texttt{devtools} package.

\begin{knitrout}\footnotesize
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{alltt}
\hlcom{# download and install helper software}
\hlkwd{install.packages}\hlstd{(}\hlstr{"devtools"}\hlstd{)}
\hlcom{# install the qlcData package from GitHub}
\hlstd{devtools}\hlopt{::}\hlkwd{install_github}\hlstd{(}\hlstr{"cysouw/qlcData"}\hlstd{,} \hlkwc{build_vignettes} \hlstd{=} \hlnum{TRUE}\hlstd{)}
\hlcom{# load the software, so it can be used}
\hlkwd{library}\hlstd{(qlcData)}
\end{alltt}
\end{kframe}
\end{knitrout}

@@

\section{Profiles and error reporting}
\label{error-reporting}

The first example of how to use these functions considers finding errors in the
encoding of texts. In the following example, it looks as if we have two
identical strings, \texttt{AABB}. However, this is just an surface-impression
delivered by the current font, which renders Latin and Cyrillic capitals
identically. We can identify this problem when we produce an orthography profile
from the strings. Using here the R implementation of orthography profiles, we
first assign the two strings to a variable \texttt{test}, and then produce an
orthography profile with the function \texttt{write.profile}. As it turns out,
some of the letters are Cyrillic.

@@

\noindent The function of error-message reporting can also nicely be illustrated
with this example. Suppose we made an orthography profile with just the two
Latin letters <A> and <B> as possible graphemes, then this profile would not be
sufficient to tokenize the strings. There are graphemes in the data that are not
in the profile, so the tokenization produces an error, which can be used to fix
the encoding (or the profile). In the example below, we can see that the
Cyrillic encoding is found in the second string of the \texttt{test} input.

@@

The function \texttt{write.profile} can be used to prepare a skeleton for an
orthography profile from some data. The preparation of an orthography profile
from some data might sound like a trivial problem, but actually there are
various different ways in which strings can be separated into graphemes by
\texttt{write.profile}. Consider the following string of characters called
\texttt{example} below. The default settings of \texttt{write.profile} separates
the string into Unicode graphemes according to grapheme clusters
(`user-perceived characters', see Section~\ref{the-unicode-approach}). As it 
turns out, ome of these graphemes are single codepoints, others are combinations
of two codepoints (see Section~\ref{pitfall-characters-are-not-glyphs}).

@@

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Thu Aug 18 14:19:34 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 1 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ú & 1 & U+0055, U+0301 & LATIN CAPITAL LETTER U, COMBINING ACUTE ACCENT \\ 
  Ù & 1 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Ù & 1 & U+0055, U+0300 & LATIN CAPITAL LETTER U, COMBINING GRAVE ACCENT \\ 
  Û & 1 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
  Û & 1 & U+0055, U+0302 & LATIN CAPITAL LETTER U, COMBINING CIRCUMFLEX ACCENT \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 1 (default settings, splitting grapheme clusters)} 
\label{tab:profile1}
\end{table}

\noindent By specifying the splitting separator as the empty string
\texttt{sep~=~""}, it is possible to split the string into Unicode codepoints,
thus separating the combining diacritics. The idea behind this option
\texttt{sep} is that separating by a character allows for user-determined
separation. The most extreme choice here is the empty string \texttt{sep~=~""},
which is interpreted as separation everywhere. The other extreme is the default
setting \texttt{sep~=~NULL}, which means that the separation is not
user-defined, but relegated to the Unicode grapheme definitions.

@@

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Thu Aug 18 14:19:34 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
́ & 1 & U+0301 & COMBINING ACUTE ACCENT \\ 
  ̀ & 1 & U+0300 & COMBINING GRAVE ACCENT \\ 
  ̂ & 1 & U+0302 & COMBINING CIRCUMFLEX ACCENT \\ 
  U & 3 & U+0055 & LATIN CAPITAL LETTER U \\ 
  Ú & 1 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 1 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 1 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 2 (splitting by codepoints)} 
\label{tab:profile2}
\end{table}

\noindent Some characters look identical, although they are encoded differently.
Unicode offers different ways of normalization (see
Section~\ref{pitfall-canonical-equivalence}), which can be invoked here as well
using the option \texttt{normalize}. NFC-normalization turns everything into the
precomposed characters, while NFD-normalization separates everything into base
characters with combining diacritics. Splitting by codepoints (i.e. \texttt{sep~=~""}) shows the results of these two normalizations.

@@

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Thu Aug 18 14:19:34 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 2 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 2 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 3 (splitting by NFC codepoints)} 
\label{tab:profile3}
\end{table}

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Thu Aug 18 14:19:34 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & Unicode Name \\ 
  \midrule
́ & 2 & U+0301 & COMBINING ACUTE ACCENT \\ 
  ̀ & 2 & U+0300 & COMBINING GRAVE ACCENT \\ 
  ̂ & 2 & U+0302 & COMBINING CIRCUMFLEX ACCENT \\ 
  U & 6 & U+0055 & LATIN CAPITAL LETTER U \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 4 (splitting by NFD codepoints)} 
\label{tab:profile4}
\end{table}

\noindent It is important to realize that for Unicode grapheme definitions, NFC
and NFD normalization are equivalent. This can be shown by normalizing the
example in either NFC or NFD, but using default separation in
\texttt{write.profile}. To be precise, default separation means setting
\texttt{sep~=~NULL}, but that has not be added explicitly.

@@

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Thu Aug 18 14:19:34 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+0055, U+0301 & LATIN CAPITAL LETTER U, COMBINING ACUTE ACCENT \\ 
  Ù & 2 & U+0055, U+0300 & LATIN CAPITAL LETTER U, COMBINING GRAVE ACCENT \\ 
  Û & 2 & U+0055, U+0302 & LATIN CAPITAL LETTER U, COMBINING CIRCUMFLEX ACCENT \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 5 (splitting by graphemes after NFD)} 
\label{tab:profile5}
\end{table}

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Thu Aug 18 14:19:34 2016
\begin{table}[H]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Gr. & Freq. & Codepoint & Unicode Name \\ 
  \midrule
Ú & 2 & U+00DA & LATIN CAPITAL LETTER U WITH ACUTE \\ 
  Ù & 2 & U+00D9 & LATIN CAPITAL LETTER U WITH GRAVE \\ 
  Û & 2 & U+00DB & LATIN CAPITAL LETTER U WITH CIRCUMFLEX \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Profile 6 (splitting by graphemes after NFC)} 
\label{tab:profile6}
\end{table}

\noindent Note that these different profiles can also be produced using the bash
executable \texttt{writeprofile} (see \ref{available-implementations} for notes
on how to install the bash executable). Exactly this example is also included in
the help file of the executable.

@@

\begin{verbatim}
## USAGE: 
##   writeprofile [-hiec] [-s SEP] [-n NORMALIZE] [<STRINGS>]
## 
## DESCRIPTION:
##   Using the function write.profile() from the R-package qlcData to prepare 
##   a skeleton for an orthography profile to be used with tokenize. Details:
##   http://www.rdocumentation.org/packages/qlcData/functions/readwrite_orthography
## 
##   STRINGS can be piped through. The results is a tab-delimited file for manual
##   editing. Note that the R-code allows for even more options, not all options 
##   are made available here for ease of use.
## 
## OPTIONS:
##   -h, --help        Showing this help text
##   -i, --info        Add columns with Unicode information to profile
##   -e, --editing     Add empty columns for easy editing of profile
##   -c, --codepoints  Separate by Unicode codepoints (overrides any -s setting)
##   -s SEP            Separator used to separate the strings
##                       defaults to unicode character definition [default: NULL]
##   -n NORMALIZE      Normalization used to make profile. Possible options: NFD, NFC
##                       defaults to no normalization [default: NULL]
##                     
## EXAMPLES:
##   Note the differences between the following variants:
## 
##     example='ÙÚÛÙÚÛ'
##     echo -e $example
##     echo -e $example | writeprofile -i
##     echo -e $example | writeprofile -ic
##     echo -e $example | writeprofile -ic -n NFC
##     echo -e $example | writeprofile -ic -n NFD
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

% latex table generated in R 3.3.0 by xtable 1.8-2 package
% Thu Aug 18 14:19:34 2016
\begin{table}[htb]
\centering
\begingroup\scriptsize
\begin{tabular}{llll}
  \toprule
Grapheme & Frequency & Codepoint & UnicodeName \\ 
  \midrule
sh &  &  &  \\ 
  ch &  &  &  \\ 
  sch &  &  &  \\ 
  ie &  &  &  \\ 
  oo &  &  &  \\ 
  a & 2 & U+0061 & LATIN SMALL LETTER A \\ 
  c & 2 & U+0063 & LATIN SMALL LETTER C \\ 
  e & 3 & U+0065 & LATIN SMALL LETTER E \\ 
  f & 1 & U+0066 & LATIN SMALL LETTER F \\ 
  h & 4 & U+0068 & LATIN SMALL LETTER H \\ 
  i & 3 & U+0069 & LATIN SMALL LETTER I \\ 
  m & 4 & U+006D & LATIN SMALL LETTER M \\ 
  o & 2 & U+006F & LATIN SMALL LETTER O \\ 
  p & 2 & U+0070 & LATIN SMALL LETTER P \\ 
  s & 4 & U+0073 & LATIN SMALL LETTER S \\ 
   \bottomrule
\end{tabular}
\endgroup
\caption{Manually edited profile skeleton} 
\label{tab:profile_skeleton1}
\end{table}

\noindent It is also possible to leave out the Unicode information in
the profile skeleton by using the option \texttt{info = FALSE}. It is also
possible not to use a separate file at all, but process everything within R. In
simply situations this is often useful (see below), but in general we prefer to
handle everything through a separately saved orthography profile. This profile
often contains highly useful information that is nicely coded and saved inside
this one file, and can thus be easily distributed and shared. Doing the same as
above completely within R might look as follows:

@@

@@

Not yet everything is correct in this graphemic parsing of the example discussed
previously. The sequence <sh> in `mishap' should not be a digraph, and
conversely the sequence <sch> in `mischief' should of course be separated into
<s> and <ch>. One of the important issues to get the graphemic parsing right is
the order in which graphemes are parsed. For example, currently the grapheme
<sch> is parsed before the grapheme <ch>, leading to <m\ i\ sch\ ie\ f> instead
of the intended <m\ i\ s\ ch\ ie\ f>. The reason that <sch> is parsed before
<ch> is that by default longer graphemes are parsed before shorter ones. Our
experience is that in most cases this is expected behaviour. You can change the
ordering by specifying the option \texttt{ordering}. Setting this option to
\texttt{NULL} results in no preferential ordering, i.e. the graphemes are parsed
in the order of the profile, from top to bottom. Now `mischief' is parsed
correctly, but `scheme' is wrong. So this ordering is not the solution in this
case.

@@

   \item \textsc{size}\\
         This option orders the lines in the profile by the size of the
         grapheme, largest first. Size is measured by number of Unicode
         characters after normalization as specified in the option
         \texttt{normalize}. For example, <é> has a size of 1 with
         \texttt{normalize = "NFC"}, but a size of 2 with
         \texttt{normalize = "NFD"}.

   \item \textsc{context}\\ This option orders the lines by whether they have
           any context specified (see next section). Lines with context will
           then be used first. Note that this only works when the option
           \texttt{regex = TRUE} is also chosen (otherwise context
           specifications are not used).

   \item \textsc{reverse}\\ This option orders the lines from bottom to top.
         Reversing order can be useful because hand-written profiles tend to put
         general rules before specific rules, which mostly should be applied in
         reverse order.

  \item \textsc{frequency}\\
         This option orders the lines by the frequency with which they
         match in the specified strings before tokenization, least frequent
         coming first. This frequency of course depends crucially on the
         available strings, so it will lead to different orderings when applied
         to different data. Also note that this frequency is (necessarily)
         measured before graphemes are identified, so these ordering frequencies
         are not the same as the final frequencies shown in the output.
         Frequency of course also strongly differs on whether context is used
         for the matching through \texttt{regex = TRUE}.

@@

By specifying more than one ordering, these orderings are used to break ties,
e.g. the default setting \texttt{ordering = c("size", "context", "reverse")}
will first order by size, and for those with the same size, it will order by
whether there is any context specified or not. For lines that are still tied
(i.e. the have the same size and both/neither have context) the order will be
reversed compared to the order as attested in the profile, because most
hand-written specifications of graphemes will first write the general rule,
followed by more specific regularities. To get the right tokenization, these 
rules should in most cases be applied in reverse order.

@@

\noindent This will lead to the following four files being written. Crucially, a
new profile is produced with the re-ordered orthography profile. To reproduce
the tokenization, this re-ordered profile can be used with the option
\texttt{ordering~=~NULL}.

   \item \textsc{result\_strings.tsv}:\\ A tab-separated file with the original
         and the tokenized/transliterated strings

   \item \textsc{result\_profile.tsv}:\\ A tab-separated file with the
         graphemes with added frequencies of occurence in the data. The lines
         in the file is re-ordered according to the order that resulted from the
         ordering specifications (see Section~\ref{rule-ordering}).

   \item \textsc{result\_errors.tsv}:\\ A tab-separated file with all original
         strings that contain unmatched parts. Unmatched parts are indicated
         with the character as specified with the option \texttt{missing}. By
         default the character \textsc{double question mark} <⁇> at
         \uni{2047} is used. When there are no errors, then this file is 
         absent.

    \item \textsc{result\_missing.tsv}:\\ A tab-separated file with the graphemes
          that are missing from the original orthography profile, as indicated in
          the errors. When there are no errors, then this file is absent.
